---
title: 'Class 2: Introduction to ggplot2'
author: Andrew Parnell \newline \texttt{andrew.parnell@mu.ie}   \newline \vspace{1cm}
  \newline \includegraphics[width=3cm]{maynooth_uni_logo.jpg}
  \newline \vspace{1cm}
  \newline PRESS RECORD 
  https://andrewcparnell.github.io/dataviz_course 
output:
  beamer_presentation:
    includes:
      in_header: header.tex
classoption: "aspectratio=169"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'pdf')
par(mar=c(3,3,2,1), mgp=c(2,.7,0), tck=-.01,las=1)
library(tidyverse)
```

## Learning outcomes

- Learn the basics of how to use `ggplot2`
- Be able to add simple features to existing ggplots
- Be able to perform basic customisation of ggplots

## The philosophy behind the grammar of graphics

- The `gg` in `ggplot2` stands for grammar of graphics. The idea is that to build a graphic we need a good grammar, just like we need grammar to write sentences
- The graphical grammar is built from geometric objects, scales and a coordinate system which are layered on top of each other
- Extra layers can be added which might alter the scales, split the plot into multiple panels, change colours, etc
- `ggplot2` requires just a few of these components to be specified, and then cleverly works out from your data what appropriate values should be used to create the plot

## Reminder: the penguin data

```{r}
library(palmerpenguins)
penguins %>% glimpse
```


## Data formats, aesthetics, geoms

Every ggplot has:

- A data set (usually a data frame or a tibble)
- An __(aes)thetic__ which maps the data to graphical elements
- A __(geom)etry__ which types of graphical elements to display

\begin{center}
\includegraphics[width=8cm]{aes.png}
\end{center}

(From [Fundamentals of Data Visualisation](https://clauswilke.com/dataviz))

## An example ggplot

```{r}
ggplot(data = penguins, 
       aes(x = bill_length_mm, y = flipper_length_mm)) + 
  geom_point()
```

(Why is this better than the plot from class 1?)

## Some basic plot types: 1 Bar charts

```{r}
ggplot(penguins, aes(x = island)) + 
  geom_bar()
```

## Some basic plot types: 2 Histograms

```{r}
ggplot(penguins, aes(x = flipper_length_mm)) + 
  geom_histogram(bins = 30)
```

## Some basic plot types: 3 Boxplots

```{r}
ggplot(penguins, aes(x = sex, y = bill_length_mm)) + 
  geom_boxplot()
```

## Some basic plot types: 4 Scatter plots

```{r}
ggplot(data = penguins, 
       aes(x = bill_length_mm, y = body_mass_g)) + 
  geom_point()
```

# Basic plot customisation and themes

## Labels and titles

Use `labs`:

```{r, eval = FALSE}
ggplot(data = penguins, 
       aes(x = bill_length_mm, y = body_mass_g)) + 
  geom_point() + 
  labs(
    title = "Bill length vs body mass",
    subtitle = "Penguin data set",
    x = "Bill length (mm)",
    y = "Body mass (g)",
  )
```

or you can specify these individually with, e.g. `+ xlab("Bill length")`

## Adding colour

Add colour as another aesthetic:

```{r}
ggplot(data = penguins, 
       aes(x = bill_length_mm, y = body_mass_g, colour = species)) + 
  geom_point()
```

(see later for manually specifying colour palettes)

## Changing point types

```{r}
ggplot(data = penguins, 
       aes(x = bill_length_mm, y = body_mass_g, 
           shape = species, size = species)) + 
  geom_point()
```

## Changing the coordinates

```{r}
ggplot(data = penguins, 
       aes(x = bill_length_mm, y = flipper_length_mm)) + 
  geom_point() + 
  coord_fixed()
```


## Some notes

- The aesthetic can go inside the `ggplot` or inside the `geom`
- If you put it inside the `ggplot` function it will persist across later layers
- By contrast if you put it inside the geom it only works for that layer

## Exercise

Every body spend 2 minutes taking this plot command and adding basic customisation to it

```{r}
ggplot(data = penguins, 
       aes(x = sex, y = flipper_length_mm)) + 
  geom_boxplot()
```



## Summary

- Most of the hard work with `ggplot2` is getting your data into the right format
- Y
- Z
