---
title: 'Class 4: High dimensional visualisation'
author: Andrew Parnell \newline \texttt{andrew.parnell@mu.ie}   \newline \vspace{1cm}
  \newline \includegraphics[width=3cm]{maynooth_uni_logo.jpg}
  \newline \vspace{1cm}
  \newline PRESS RECORD 
  \newline https://andrewcparnell.github.io/dataviz_course 
output:
  beamer_presentation:
    includes:
      in_header: header.tex
classoption: "aspectratio=169"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'pdf', fig.height = 4,
                      warning=FALSE, message=FALSE)
library(tidyverse)
```

## Learning outcomes

- Learn how to create multiple scatter plots
- Plot some map data
- Plot some time series data
- Use some of the 3D geoms
- Our last session where we mostly use ggplot

## New data set for this session

Geese isotopes: Two isotopes taken from the blood plasma of geese together with information about each of the birds

\small 
```{r}
geese <- read.csv('../data/geese_isotopes.csv')
geese %>%  glimpse
```

## Multiple scatter plots: `pairs`

```{r}
pairs(geese[,1:5])
```

## ggplot version using package `GGally`

```{r, message=FALSE, warning=FALSE}
library(GGally)
ggpairs(geese[,1:5])
```

Very customisable too - can change what the upper, lower and diagonal panels contain or add colours to different groups

## Extended ggpairs plot

```{r}
ggpairs(geese[,1:5],
        upper = list(continuous = "density"),
        lower = list(continuous = "points"))
```


## Visualise correlations

```{r, fig.height=2.5, fig.align='center'}
library(corrplot)
corrplot(cor(geese)) # Note: need a correlation matrix!
```

## 2D density plots

`ggplot2` has specific functions for plotting 3D density data

```{r, fig.height=2.5}
ggplot(geese, aes(x = d15NPl, y = d13CPl)) + 
  geom_bin_2d() + theme(aspect.ratio = 1)
```

- There are related functions `geom_tile` and `geom_raster` when you have a variable by which you want to colour the pixels. 

## Contour plots

```{r,fig.height=3.5}
ggplot(geese, aes(x = d15NPl, y = d13CPl)) + 
  geom_density_2d_filled() + theme(aspect.ratio = 1)
```

- Again, use `geom_contour` if you have a variable to directly create the contour values

## Hexagonal plots

```{r,fig.height=3.5}
ggplot(geese, aes(x = d15NPl, y = d13CPl)) + 
  geom_hex() + theme(aspect.ratio = 1)
```

## Plotting data with maps

- Most base R plots of maps are ugly. `ggplot` has an interface to the `maps` package but these are quite ugly and out of date
- Better package are `ggmap`, `ggspatial` and `sf` (simple features) which have loads of examples and interfaces for both base R and ggplots
- Three common use cases:

  - You want to take a map of an area and add information to it by filling in the areas or placing points on the map
  - You want to take an e.g. Google or OpenStreetMap (raster) map and add information on top
  - You have your own shape file and want to create a plot on top of that (not covered -- see `read_sf`)
  
## Scenario 1: adding information to a map of a given area

```{r, fig.show='hide'}
library(maps)
norway <- map_data("world", "norway")
norway %>% glimpse
ggplot(norway, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "grey50") + 
  coord_quickmap()
```

## Scenario 1: plot

```{r, echo = FALSE, fig.align='center', fig.height=3.5}
ggplot(norway, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "grey50") + 
  coord_quickmap()
```

The `group` variable is necessary so that it joins the islands correctly

## Adding information

```{r, fig.show='hide'}
points <- tibble(
  long = c(0, 10, 20),
  lat = c(65, 70, 75), 
  group = 1 # Necessary to match the fitting command
)
ggplot(norway, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "grey50") + 
  coord_quickmap() + 
  geom_point(data = points) # Add colours etc here
```

## Plot with added information

```{r, echo = FALSE, fig.align='center'}
ggplot(norway, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "grey50") + 
  coord_quickmap() + 
  geom_point(data = points) # Add colours etc here
```

## Scenario 1 (cont): with e.g. contours

```{r, fig.show='hide'}
new_data <- MASS::mvrnorm(n = 100, 
                          mu = c(10, 70), 
                          Sigma = diag(10, 2)) %>% 
  as_tibble() %>% 
  rename(long = V1, lat = V2) %>% 
  mutate(group = 1)
ggplot(norway, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "grey50") + 
  coord_quickmap() + 
  geom_density_2d_filled(data = new_data, alpha = 0.5)
```

## Scenario 1: plot

```{r, echo = FALSE, fig.align='center'}
ggplot(norway, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "grey50") + 
  coord_quickmap() + 
  geom_density_2d_filled(data = new_data, alpha = 0.5)
```


## Scenario 2: overlaying an existing web map

```{r, fig.show='hide'}
library(ggspatial)
ggplot(new_data, aes(x = long, y = lat)) + 
  annotation_map_tile(type = "osm") +
  geom_density_2d_filled(alpha = 0.5) + 
  coord_sf(crs = 4326)  
```

## Scenario 2: plot

```{r, echo = FALSE, fig.align='center', fig.height = 3}
ggplot(new_data, aes(x = long, y = lat)) + 
  annotation_map_tile(type = "osm") +
  geom_density_2d_filled(alpha = 0.5) + 
  coord_sf(crs = 4326)  
```

## Leaflet

See demo in Rstudio:

```{r, eval = FALSE}
library(leaflet)
leaflet() %>%
  addTiles() %>% 
  setView(lng = 10, lat = 62, zoom = 5) %>% 
  addMarkers(lng = 10, lat = 62) %>% 
  addCircles(lng = 11, lat = 63, radius = 5000)
```


## Common problems with chloropleth maps

\begin{center}
\includegraphics[width=10cm]{election_map.png}
\end{center}

From [Wikipedia](https://en.wikipedia.org/wiki/Choropleth_map#/media/File:U.S._Presidential_election_margin,_2004-2016.png)

## Plotting time series data

Sea level data set:

```{r}
sl_month <- read.csv('../data/sea_level_monthly.csv')
glimpse(sl_month)
```

## Plots of time series

```{r, fig.height=3}
sl_month %>% 
  mutate(Year2 = lubridate::date_decimal(Year)) %>%
  ggplot(aes(x = Year2, y = GMSL_mm)) + 
  geom_line() + 
  scale_x_datetime(date_breaks = "20 years",
                   date_labels = "%Y-%m")
```

## dygraphs - interactive time series

See Rstudio demo

```{r, eval = FALSE}
library(dygraphs)
sl_month %>% 
  select(Year, GMSL_mm) %>% 
  dygraph() %>% 
  dyRangeSelector()
```

## Automatic plotting with `ggfortify`

 - There is a function in `ggplot2` called `autoplot` that tries to take into account the object class when creating a plot
 - Other packages can use and extend this function, for example the `forecast` package uses `autoplot` to create nice time series plots
 - The `ggfortify` package uses this even more generally to create nice plots for a load of different classes of object
 
## Principal Components Analysis and `ggfortify`

```{r, fig.height = 3.5}
library(ggfortify)
pollen <- read.csv('../data/pollen.csv')
pca <- prcomp(pollen[,3:9], scale. = TRUE)
autoplot(pca, data = pollen, colour = "MTCO",
         loadings = TRUE, loadings.label = TRUE)
```

## Cluster analysis and `ggfortify`

```{r}
autoplot(kmeans(pollen[,3:9], 3), 
         data = pollen)
```

- Can also do generalised linear models, maps, time series, other dimension reduction methods (factor analysis and multidimensional scaling), and survival data

## Exercise

Run some of the above dimension reduction techniques on the `geese` data above and produce some plots.  Try removing different columns (e.g. the time columns. What does the PCA tell you? (Post plots and answers in Slack)

```{r, include = FALSE}
pca <- prcomp(geese[,2:5], scale. = TRUE) # Or remove 2:5
autoplot(pca, data = geese,
         loadings = TRUE, 
         loadings.label = TRUE)

```

## Summary

- Lots of options for high dimensional data. Start with matrix scatter plots
- Plotting maps is a minefield. Pick a package and work through it for what you want to produce. There are many other packages and complications
- Always a good idea to try some dimension reduction when plotting high dimensional data; `ggfortify` a simple first choice
